UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts

### 2019-06-03 ms/main.tex
moar tweaks

Focusing on UMAP with 15 principal components (figure~\ref{fig:1000g_compare}~(iv)), several population clusters reflect shared ancestries. British individuals from England and Scotland form a cluster mixed with those from Utah who claim Northern and Western European ancestry. Toscani and Iberian individuals form a group reflecting their Mediterranean heritage. African Americans in the Southwest US, African Caribbean individuals in Barbados, and some Puerto Ricans also form a cluster. Three East Asian groups appear: one is largely Han and Southern Han individuals, another is comprised of the Chinese Dai in southern China and the Kinh from Vietnam, and the third is the Japanese population. Looser geographical groupings include Colombians and Peruvians, and the Esan and Yoruba populations of Nigeria; both groupings appear as connected sub-clusters. The South Asian populations also form a loose grouping.

\adpcomment{trim this pgh?}
Only a few individuals cluster differently than the majority of individual bearing the same population label: a few Mexican individuals cluster with Spanish and Italian individuals, and a few Puerto Ricans cluster with the African Americans and African Caribbeans, likely resulting from ancestry proportions that differ from the majority. One Gambian-identified individual is present in a cluster that is otherwise entirely Mende people from Sierra Leone. Two populations form multiple clusters: Gujarati Indians in Houston, Texas and Punjabi people in Lahore, Pakistan. This clustering effect is robust to the number of components considered (see figure~\ref{fig:supp_megamontage_pc10_50}). Differentiation in the Gujarati population from substructure has been previously noted\cite{reich2009india}. Recontacting of 23andMe participants associated with the clusters show that one of the cluster is highly enriched in individuals from a group in Western India with shared ancestry and patronym. [D. Poznik, 23andme, personal communication, and \cite{23andme}].

\subsubsection*{The genetic continuum of admixed populations}
The 1KGP sampled individuals from relatively distinct populations, so the data are more likely to form clusters. Most medical cohorts, however, comprise larger numbers of individuals sampled across extended geographical areas. The HRS contains genotype data of 12,454 Americans from a variety of backgrounds. Using UMAP on the first 10 principal components, we demonstrate projections that present a collection of sub-populations and a continuum of genetic variation.

>>>>

Focusing on UMAP with 15 principal components (figure~\ref{fig:1000g_compare}~(iv)), several population clusters reflect shared ancestries. British individuals from England and Scotland form a cluster mixed with those from Utah who claim Northern and Western European ancestry. Toscani and Iberian individuals form a cluster reflecting their Mediterranean heritage. African Americans in the Southwest US, African Caribbean individuals in Barbados, and some Puerto Ricans also form a cluster. Three East Asian clusters appear: one is largely Han and Southern Han individuals, another is comprised of the Chinese Dai in southern China and the Kinh from Vietnam, and the third is the Japanese population. Other clusters are comprised of Colombians and Peruvians, the Esan and Yoruba populations of Nigeria, and several South Asian populations.

\adpcomment{changed this a bit. or maybe remove most of it?}
A few individuals cluster with populations different from their label: some Mexican individuals cluster with Spanish and Italian populations; some Puerto Rican individuals cluster with African American and Caribbean populations; and one Gambian individual clusters with the Mende of Sierra Leone. Two populations form multiple clusters: Gujarati Indians in Houston, Texas and Punjabi people in Lahore, Pakistan. This clustering effect is robust to the number of components considered (see figure~\ref{fig:supp_megamontage_pc10_50}). Differentiation in the Gujarati population from substructure has been previously noted\cite{reich2009india}. Recontacting of 23andMe participants associated with the clusters show that one of the cluster is highly enriched in individuals from a group in Western India with shared ancestry and patronym. [D. Poznik, 23andme, personal communication, and \cite{23andme}].

%of individual bearing the same population label: a few Mexican individuals cluster with Spanish and Italian individuals, and a few Puerto Ricans cluster with the African Americans and African Caribbeans, likely resulting from ancestry proportions that differ from the majority. One Gambian-identified individual is present in a cluster that is otherwise entirely Mende people from Sierra Leone. Two populations form multiple clusters: Gujarati Indians in Houston, Texas and Punjabi people in Lahore, Pakistan. This clustering effect is robust to the number of components considered (see figure~\ref{fig:supp_megamontage_pc10_50}). Differentiation in the Gujarati population from substructure has been previously noted\cite{reich2009india}. Recontacting of 23andMe participants associated with the clusters show that one of the cluster is highly enriched in individuals from a group in Western India with shared ancestry and patronym. [D. Poznik, 23andme, personal communication, and \cite{23andme}].

%The genetic continuum of admixed populations
\adpcomment{cleaned up, changed subtitle}
\subsubsection*{Admixed individuals fall along a genetic continuum}
The 1KGP sampled individuals from relatively distinct populations, so the data are more likely to form clusters. Most medical cohorts, however, comprise larger numbers of individuals sampled across extended geographical areas. The HRS contains genotype data of 12,454 Americans from a variety of backgrounds. Using UMAP on the first 10 principal components, we demonstrate projections that present a collection of sub-populations and a continuum of genetic variation.
